---
title: "Human Activity Recognition"
author: "Izaak Jephson"
date: "11/08/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("ggplot2")
library("caret")
library("lubridate")
library("rattle")
library("doParallel")
library("randomForest")
library("scales")
library("corrplot")
```

## Introduction

## Data
This data set was provided by Velloso, E., Bulling, A., Gellersen, H., Ugulino, W., Fuks, H.

In this study, six participants were asked to perform one set of 10 repetitions of the unilateral dumbbell biceps curl in five different ways, according to a  specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).


```{r read}

## Read in and tidy data

rawData <- read.csv(file = "data/pml-training.csv") 

data <- rawData%>%
        as_tibble() %>% 
        mutate_all(
                .funs = function(x)
                        replace(x, which(x == "" | x == "#DIV/0!"), NA) # manually replace missing values to ensure no useful data lost in tidying
        ) %>%
        mutate_at(vars(
                -c(X,
                   user_name,
                   cvtd_timestamp,
                   new_window,
                   num_window,
                   classe) 
        ), as.numeric) %>% # convert features to numeric variables
        mutate_at("classe", as_factor) %>% 
        mutate_at("cvtd_timestamp", dmy_hm) %>% 
        select_if(function(x) (mean(is.na(x))) < 0.9) %>% # remove rows with no data
        select(-c("user_name",
                  "X",
                  "raw_timestamp_part_1",
                  "raw_timestamp_part_2",
                  "cvtd_timestamp",
                  "new_window",
                  "num_window"))

rawFinalTest <- read.csv(file = "data/pml-testing.csv")

finalTest <- rawFinalTest %>% # apply same tidying to testing set
        as_tibble() %>% 
        mutate_all(
                .funs = function(x)
                        replace(x, which(x == "" | x == "#DIV/0!"), NA) # manually replace missing values to ensure no useful data lost in tidying
        ) %>%
        mutate_at(vars(
                -c(X,
                   user_name,
                   cvtd_timestamp,
                   new_window,
                   num_window,
                   problem_id) 
        ), as.numeric) %>% # convert features to numeric variables
        mutate_at("cvtd_timestamp", dmy_hm) %>% 
        select_if(function(x) (mean(is.na(x))) < 0.9)

```

## Inital data exploration

In our data, there are `r data %>% nrow()` observations and `r data %>% ncol()` variables. the variable that we are interested in predicting is the "classe" variable, which is a factor variable with 5 levels (A, B, C, D, E), each of which represents a different way in which the exercise was performed. 

```{r slice}

# Create data partition and split data into training and test data in order to be able to estimate out of sample error

inTrain <- data %>% pull(classe) %>% createDataPartition(p = 0.75, list = FALSE) 

training <- data %>% slice(inTrain)
testing <- data %>% slice(-inTrain)

```


```{r correlation}

training %>% select(-classe) %>% cor() %>% corrplot(method = "color",type = "upper", order = "FPC", tl.cex = 0.5)

trainingPC <- training %>% select(-classe) %>% preProcess(method = "pca", pcaComp = 10) %>% predict(training %>% select(-classe))

trainingPC %>% cor() %>% corrplot(method = "color",type = "upper", order = "FPC", tl.cex = 0.5)


```


```{r tree}

# Fit a decision tree to the data 

set.seed(2572)
model1 <- train(classe ~ ., data = training, na.action = na.pass, method = "rpart", trControl = trainControl(method = "cv", number = 10))
fancyRpartPlot(model1$finalModel, cex = 0.6, sub = "")

predict1 <- predict(model1, testing)

confMat1 <- confusionMatrix(testing %>% pull(classe), predict1)

```

The accuracy of the decision tree model on the test set is `r confMat1$overall["Accuracy"] %>% percent()` and the out of sample error rate is `r (1 - confMat1$overall["Accuracy"]) %>% percent()` 


```{r forest, cache = TRUE}

set.seed(591)

model2 <- train(classe ~ ., data = training, na.action = na.pass, method = "rf", trControl = trainControl(method = "cv", number = 10))

predict2 <- predict(model2, testing)

confMat2 <- confusionMatrix(testing %>% pull(classe), predict2)


```

The accuracy of the random forest model on the test set is `r confMat2$overall["Accuracy"] %>% percent()` and the out of sample error rate is `r (1 - confMat2$overall["Accuracy"]) %>% percent()` 

```{r regression}

set.seed(70333)

model3 <- train(classe ~ ., data = training, na.action = na.omit, method = "treebag", trControl = trainControl(method = "cv", number = 10))

predict3 <- predict(model3, testing)

confMat3 <- confusionMatrix(testing %>% pull(classe), predict3)


```


The accuracy of the bagged tree model on the test set is `r confMat3$overall["Accuracy"] %>% percent()` and the out of sample error rate is `r (1 - confMat3$overall["Accuracy"]) %>% percent()` 

## References

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.

Read more: http:/groupware.les.inf.puc-rio.br/har#ixzz4Tk2YB4Ft

